Computational biology in the 21st century:\\Scaling with compressive algorithms

### 2015-09-21 main.tex
Hopefully ready to submit.

\textbf{chemogenomics} This means this and that

\textbf{metagenomics} This means another thing and so on and so forth blah blah blah blah blah.

\textbf{metabolomics} And more.

\textbf{shotgun sequencing} And more.

\textbf{homology search} Determining the function, structure, or identity of a gene sequence by locating similar sequences within an annotated database.

\textbf{transcriptome} And more.

\textbf{FASTQ} A common plain-text format for biological sequence data.

\textbf{BLAST} And more.

@@

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into polypeptide chains, sequences of amino 
acids which singly or in complexes are known as proteins.
Proteins fold into complex, low-energy structures, which function as
cellular machines; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of a protein.
This structure ultimately determines a protein's function within the cell.
Certain kinds of RNA also function, sometimes with proteins, as cellular 
machines.
Methods have been developed to gather biological data from every level of this process,
resulting in a massive influx of data on sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins,
Much of these data are amenable to standard Big Data analysis methods; however, in this Review we focus on several examples of biological data which exhibit additional exploitable structure for creating scalable algorithms.

Sequence data, either 
nucleotide sequences (using a four-letter alphabet representing the four
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids)
are obtained in several ways.
For protein and RNA sequence data, mass spectrometry---which
can determine protein sequence, as well as interactions---and RNA-seq---which can determine RNA sequence and abundance, 
allowing scientists to also infer the expression of the gene to which it might translate---play central roles.
However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
To better understand the structure of NGS sequence data, we expand a little on NGS methodologies.

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
Though slow and labor intensive, Sanger sequencing has the advantage of being
well-tested and is still used to validate NGS results.
More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's `sequencing by synthesis,' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, in so-called `shotgun sequencing' DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads must be put together in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
Because no sequencing technology is completely infallible, sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many overlapping reads, typically stored in a FASTQ
file.

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease, and 
obesity~\cite{kostic2014microbiome}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

The computational study of drugs and their targets based on chemical structure and function
is known as chemogenomics~\cite{bredel2004chemogenomics}.
In the fields of drug discovery and drug repurposing, the prediction 
of biologically active compounds is an important task. 
Computational high-throughput screening eliminates many compounds from 
laborious wet-lab consideration, but even computational screening can be 
time-consuming.
Chemogenomics typically relies on comparing chemical graph structures to identify
similar molecules and binding sites.
Furthermore, comparing chemical graph structures typically involves computing
the maximal common subgraph (MCS), an NP-hard problem.
However, there are once again an increasing number of such chemical compounds
to search; the NCBI's PubChem database has grown from 31 million compounds in
January 2011 to 68 million in July, 2015.

@@

Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future.
Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression ratios.

@@

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Far fewer biological sequences exist than could be enumerated, but even more so, those that exist tend to be highly similar to many others.
Thanks to 
evolution, only those genes
which exhibit useful biological function survive, and most
random sequences of amino acids would not be expected to form stable structures.
Since two human genomes differ on average by only 0.1\%, a collection of 1,000 human genomes contain less than twice the unique information of a single genome~\cite{loh2012compressive}.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
It seems that physical laws---in this case, evolution---have constrained the data to a particular subspace of the enumerable Cartesian space.

\begin{figure}[htb!]
\centering
\includegraphics[width=2.8in]{assets/treepoints-fractal.png}
\caption{Cartoon depiction of points in an arbitrary high-dimensional space, 
as might arise from genomes generated by mutation and selection during the 
course of evolution. 
Although high-dimensional locally, at the global scale of covering spheres, the 
data cloud looks nearly 1-dimensional, which enables entropy-scaling of 
similarity search. Clusters cover the data points but do not cover unoccupied 
regions of space.
The green triangle represents a query, with two concentric search radii (red 
circles) around it. Thanks to low fractal dimension, the large circle does
not contain vastly more points than the small circle.}
\label{fig:dataspace}
\end{figure}

@@

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets. Metric entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\tabcolsep=0.11cm
\begin{tabular}{H{3.0cm}p{2.2cm}p{1.8cm}}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide~sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein~sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein~structure (PDB) & 10:1 & 2.5\\
\hline
Chemical~structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

\section{The age of compressive algorithms}\label{omics}

@@

Compressive genomics is based on the concept of \emph{compressive acceleration},
which relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any representative sequence that is within some threshold of the query is then expanded 
into all the similar sequences it represents; the fine search is then over this
(typically small) subset of the original database.
This approach provides orders-of-magnitude run-time improvements to BLAST search on 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
data; moreover, these run-time improvements increase as databases grow.

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST (compressively-accelerated BLAST)~\cite{loh2012compressive} and 
caBLASTP~\cite{daniels2013compressive}, accelerates existing tools (in this 
case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
As an aside, though not an example of accelerated similarity search, mid-size $l$-mers
can also be used for storage compression of NGS quality scores associated with base calls.
Unlike many other approaches to compressing quality scores in the literature
~\cite{bonfield2013compression,ochoa2013qualcomp},
Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
Because Quartz injects information from the distribution of $l$-mers in the target genome in this lossy compression, it demonstrates not only improved compression over competing
approaches, but slightly improves the accuracy of downstream variant-calling.

@@

The compressive omics approach can, in some cases, come at the cost of accuracy.
However, these cases are well-defined.
Compressive omics never results in false positives (with respect to the na\"ive 
search technique being accelerated), because the fine search phase applies the
same comparison to the candidates as the na\"ive approach.
Furthermore, when the distance function used for comparisons is a metric---more specifically, when it obeys the triangle inequality---false negatives will also
never occur.
However, in practice, non-metric distance functions are used, such as E-values
in BLAST or cosine distance in esFragBag, and thus false negatives can occur.
In practice, these error rates are low, and recall better than 90\% has been
demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}

\section{Conclusions and future prospects}

@@

The field of computational biology must continue to innovate, but also to 
incorporate the best ideas from other areas of computer science.
For example, the compressive-acceleration approach bears similarity to a metric 
ball tree, first described in the database community over twenty years 
ago~\cite{uhlmann1991satisfying};
however, the latter does not allow one to analyze performance guarantees, 
in terms of metric entropy and fractal dimension.
Other ideas from image processing, computational geometry~\cite{indyk1998approximate}, sublinear-time algorithms~\cite{rubinfeld2011sublinear}, and other areas 
outside of biology are likely to bear fruit. 
In turn, it is also likely that algorithmic ideas developed within computational biology
will become useful in other fields experiencing a data deluge, such as astronomy
or social networks~\cite{stephens2015big}.

>>>>

\textbf{chemogenomics} Computational study of drugs and their targets based on chemical structure and function

\textbf{metagenomics} Study of the many genomes that make up a particular environment

\textbf{shotgun sequencing} Modern genomic sequencing, which chops DNA into many short pieces

\textbf{homology search} Determining the function, structure, or identity of a gene sequence by locating similar sequences within an annotated database

\textbf{transcriptome} Transcribed RNA from a genome, which results in protein production

\textbf{BLAST} Standard biological sequence similarity search tool

@@

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into polypeptide chains, sequences of amino 
acids which singly or in complexes are known as proteins.
Proteins fold into complex, low-energy structures, which function as
cellular machines; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of a protein.
This structure ultimately determines a protein's function within the cell.
Certain kinds of RNA also function as cellular 
machines.
Methods have been developed to gather biological data from every level of this process,
resulting in a massive influx of data on sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins.
Much of these data are amenable to standard Big Data analysis methods; however, in this Review we focus on examples of biological data which exhibit additional exploitable structure for creating scalable algorithms.

Sequence data, either 
nucleotide sequences (using a four-letter alphabet representing the four
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids)
are obtained in several ways.
For protein and RNA sequence data, mass spectrometry---which
can determine protein sequence and interactions---and RNA-seq---which can determine RNA sequence and abundance, 
allowing scientists to also infer the expression of the gene to which it might translate---play central roles.
However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
To better understand the structure of NGS sequence data, we will expand on NGS methodologies.

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
Though slow and labor intensive, Sanger sequencing has the advantage of being
well-tested and is still used to validate NGS results.
More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's `sequencing by synthesis,' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, in so-called `shotgun sequencing' DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads must be put together in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
Because no sequencing technology is completely infallible, sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many overlapping reads.

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease, and 
obesity.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

The computational study of drugs and their targets based on chemical structure and function
is known as chemogenomics~\cite{bredel2004chemogenomics}.
In the fields of drug discovery and drug repurposing, the prediction 
of biologically active compounds is an important task. 
Computational high-throughput screening eliminates many compounds from 
laborious wet-lab consideration, but even computational screening can be 
time-consuming.
Chemogenomics typically relies on comparing chemical graph structures to identify
similar molecules and binding sites.
Furthermore, comparing chemical graph structures typically involves computing
the maximal common subgraph (MCS), an NP-hard problem.
However, there are an increasing number of such chemical compounds
to search; the NCBI's PubChem database has grown from 31 million compounds in
January 2011 to 68 million in July, 2015.

@@

Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future.
Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression.

@@

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Far fewer biological sequences exist than could be enumerated, but even more so, those that exist tend to be highly similar to many others.
Thanks to 
evolution, only those genes
which exhibit useful biological function survive, and most
random sequences of amino acids would not be expected to form stable structures.
Since two human genomes differ on average by only 0.1\%, a collection of 1,000 human genomes contain less than twice the unique information of a single genome~\cite{loh2012compressive}.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
It seems that physical laws---in this case, evolution---constrain the data to a particular subspace of the Cartesian space.

\begin{figure}[htb!]
\centering
\includegraphics[width=2.5in]{assets/treepoints-fractal.png}
\caption{Cartoon depiction of points in an arbitrary high-dimensional space, 
as might arise from genomes generated by mutation and selection during the 
course of evolution. 
Although high-dimensional locally, at the global scale of covering spheres, the 
data cloud looks nearly 1-dimensional, which enables entropy-scaling of 
similarity search. Clusters cover the data points but do not cover unoccupied 
regions of space.
The green triangle represents a query, with two concentric search radii (red 
circles) around it. Thanks to low fractal dimension, the large circle does
not contain vastly more points than the small circle.}
\label{fig:dataspace}
\end{figure}

@@

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four data sets. Metric entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
NCBI's non-redundant `NR' protein and `NT' nucleotide sequence databases are from June, 2015. Protein Data Bank (PDB) is from July, 2015. PubChem is from October, 2013.}
\tabcolsep=0.11cm
\begin{tabular}{H{3.0cm}p{2.2cm}p{1.8cm}}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide~sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein~sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein~structure (PDB) & 10:1 & 2.5\\
\hline
Chemical~structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

\section{The age of compressive\\ algorithms}\label{omics}

@@

Compressive genomics is based on the concept of \emph{compressive acceleration},
which relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any representative sequence within some threshold of the query is then expanded 
into all similar sequences it represents; the fine search is over this
(typically small) subset of the original database.
This approach provides orders-of-magnitude run-time improvements to BLAST 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
search; these run-time improvements increase as databases grow.

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST (compressively-accelerated BLAST)~\cite{loh2012compressive} and 
caBLASTP~\cite{daniels2013compressive}, accelerates existing tools (in this 
case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
As an aside, though not an example of accelerated similarity search, mid-size $l$-mers
can also be used for storage compression of NGS quality scores associated with base calls.
Unlike many other approaches to compressing quality scores in the literature
~\cite{bonfield2013compression,ochoa2013qualcomp},
Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing
approaches, but slightly improves the accuracy of downstream variant-calling.

@@

The compressive omics approach can, in some cases, come at the cost of accuracy.
However, these cases are well-defined.
Compressive omics never results in false positives (with respect to the na\"ive 
search technique being accelerated), because the fine search phase applies the
same comparison to the candidates as the na\"ive approach.
Furthermore, when the distance function used for comparisons is a metric---more specifically, when it obeys the triangle inequality---false negatives will also
never occur.
However, in practice, non-metric distance functions are used, such as E-values
in BLAST or cosine distance in esFragBag, and thus false negatives can occur.
In practice, these error rates are low, and recall better than 90\% has been
demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}.

\section{Conclusions}

@@

The field of computational biology must continue to innovate, but also to 
incorporate the best ideas from other areas of computer science.
For example, the compressive-acceleration approach bears similarity to a metric 
ball tree, first described in the database community over twenty years 
ago~\cite{uhlmann1991satisfying};
however, the latter does not allow one to analyze performance guarantees, 
in terms of metric entropy and fractal dimension.
Other ideas from image processing, computational geometry~\cite{indyk1998approximate}, sublinear-time algorithms~\cite{rubinfeld2011sublinear}, and other areas 
outside of biology are likely to bear fruit. 
It is also likely that algorithmic ideas developed within computational biology
will become useful in other fields experiencing a data deluge, such as astronomy
or social networks~\cite{stephens2015big}.
